Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-19 icfp.tex
typos and minor stuff

We start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f6 x y use = case (x,y) of z { (a,b) -> use a b }
\end{lstlisting}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is 
\emph{semantically} linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f7 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
We justify that \incode{f7} is linear by appealing to the fact that 
the tuple being scrutinized is already in WHNF and so evaluating it will
not consume either $x$ nor $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we are not using $a$ or $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
\begin{noway}
\begin{lstlisting}
f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f8 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable we run the computations
defined in terms of the linear variables used for that constructor
argument. Otherwise, if we do not use the pattern variables, then we
do not run the computation and thus no resources are consumed.

A third option is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $a,b$ and with $x,y$, but at least one of
the three must occur to ensure the linear resources are consumed. In this
example, we can think that using $a$ entails using the resource $x$, $b$ the
resource $y$, and the case binder $z$ entails using both $a$ and $b$.
%
Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require performing computation and thus consume linear
resources:
\begin{notyet}
\begin{lstlisting}
f10 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f10 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we must still use either the case binder $z$ or the linear
pattern variables $a,b$. %  e.g. it would be quite disastrous if any of the following typechecked:
% \begin{noway}
% \begin{lstlisting}
% doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
% doubleFree x free = case free x of z { Result v -> free x }
% \end{lstlisting}
% \end{noway}
% % Remover
% \begin{noway}
% \begin{lstlisting}
% leakPointer :: Ptr -> ()
% leakPointer x = case id x of z { _ -> () }
% \end{lstlisting}
% \end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly once to guarantee
that the resources used in the scrutinee are fully consumed. %, or
                                %risk them being only ``almost''
                                %consumed.
For instance, take
\incode{use} in \incode{f10} to be \incode{(,)}: it is not sufficient for \incode{use x y} to be
evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, potentially
discarding linear resources. In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

@@

Finally, we discuss case expressions which scrutinize a
single variable $x$, e.g.:
$
\lambda x.~\ccase{x}{\_ \to x}
$. 
It might seem as though this function is linear:
%begin{itemize}
% \item
if $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact;
% \item
if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
$x$ returns the result of evaluating $x$ that still had to be consumed exactly
once.
%\end{itemize}
However, in practice, this program's linearity can only be determined by
further branching down on the non-strict evaluation strategy. If linear
function applications, in general, are $\beta$-reduced \emph{call-by-name} (a desireable
practice, as linear functions are known to use their argument exactly once)
% and thus call-by-name reduction will definitely not duplicate computation
while the above function is considered linear, then such an application might
duplicate linear resources during evaluation. For instance:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
\Longrightarrow_{CBN}
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
On the other hand, if linear function applications are reduced
\emph{call-by-need}, the above example would instead be semantically linear
because the result of freeing $x$ would be bound to some $y$ first.
%
The type system we present in the next section, which evaluates linear
applications call-by-name, does not accept the above function scrutinizing a
variable as linear. 
This interaction between evaluation and linearity comes up in the type
preservation proof of our program, and is again explored with the reverse
binder swap transformation in
Section~\ref{sec:optimisations-preserve-types-meta}.

@@

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D_1}\s_1\dots x_n{:}_{\D_n}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. We often omit the case binder $z$ when it is unused. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for type-checking semantic linearity in the presence of laziness we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default} or \emph{wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting whether the pattern-bound variables must
consumed linearly (ultimately, in order to consume the scrutinee linearly).
Additionally, the set of patterns in a case expression is guaranteed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.

@@

\vspace{-0.5cm}
\[
\TypeVarDelta
\]
The rule reads that an occurrence of a $\Delta$-bound variable is well-typed if
the linear environment is exactly the resources in the usage environment of
that variable.
%
$\D$-variables are unrestricted 
since they can be discarded and duplicated, despite multiple
occurrences of the same $\Delta$-variable not being
well-typed as, ultimately, it would imply non-linear usage of linear
resources.
% multiple occurrences of the same
% $\D$-variable imply non-linear usage of resources that must be used linearly.

@@

Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
a case expression \emph{evaluates its scrutinee} to Weak Head Normal
Form (WHNF)~\cite{10.5555/1096899}, and \emph{then} selects the case
alternative corresponding to the pattern matching the WHNF of the
scrutinee\footnote{In our calculus, the alternatives are always
  exhaustive, i.e. there always exists at least one pattern which
  matches the scrutinee in its WHNF, so we're guaranteed to have an
  expression to progress evaluation.}. An expression in WHNF can either be %\begin{itemize}
% \item
a $\lambda$-abstraction $\lambda x.~e$ 
or a datatype constructor application $K~\ov{e}$.
%\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments need not be evaluated for the lambda or constructor
application to be in WHNF. % (if otherwise all sub-expressions
% were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
Accordingly, these sub-expressions might still depend on linear resources to be
well-typed (these resources will be consumed when the expressions are evaluated).

As will be made clear in this section, we need to devise a specialized typing
judgement for scrutinees that is able to distinguish between terms
in WHNF and terms that are not in WHNF.
Following the discussion on expressions in WHNF, we present a
typing judgement $\G;\D \Vdash e : \s \gtrdot \ov{\D_i}$, and a rule for each
of the forms given above:
\[
    \TypeWHNFCons
\qquad
    \TypeWHNFLam
\]
This judgement differs from the main typing judgement in that (1) it only
applies to expressions in weak head normal form, and (2) it ``outputs'' (to the right of $\gtrdot$) a
disjoint set of linear environments ($\ov{\D_i}$), where each environment corresponds to the
linear resources used by a sub-expression of the WHNF expression.
%
To type a constructor application $K~\ov{e_\omega e_i}$, where $e_\omega$
are the unrestricted arguments and $e_i$ the linear arguments of the
constructor, we split the resources $\D$ into a disjoint set of resources
$\ov{\D_i}$ required to type each linear argument individually and return exactly
that split of the resources; the unrestricted $e_\omega$ expressions must be
typed on an empty linear environment. A lambda expression is typed with
the main typing judgement and trivially ``outputs'' the whole $\D$ environment,
as there is always only a single sub-expression in lambda abstractions (the body).

@@

The dichotomy between evaluation (i.e.,~resource usage) of a case expression
whose scrutinee is in weak head normal form, or otherwise, leads to one of our
key insights: we must \emph{branch on weak head normal formed-ness} to
accurately type case expressions.
%
When the scrutinee is already in WHNF, the resources are
unused upon evaluation and thus available in the alternatives.
%
When it is not, resources will be consumed and cannot be used in the
alternative.
%
To illustrate, consider the following case expressions:
\[
\begin{array}{ccc}
(1)~\lambda x.~\ccase{K~x}{\_ \to x} &  & (2)~\lambda x.~\ccase{free~x}{\_ \to x}
\end{array}
\]
The first function uses $x$ linearly, while the second does not.
%
Alternatives may also use the case binder or pattern variables, referring to, respectively,
the whole scrutinee (and all resources used to type the scrutinee) or
constructor arguments (and the resources to type each argument).
% Using the case binder entails using all the resources required by the scrutinee, and using a pattern
% variable implies using the resources of the corresponding constructor argument.

@@

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, encompassing three
``modes``, distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF ($\Mapsto$),
for case expressions in which the scrutinee is not in WHNF ($\Rrightarrow$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t
The rule for typing an alternative whose pattern is a
constructor with $n > 0$ linear components (in mode $\Mapsto$) is:
\[
\TypeAltNWHNF
\]
The rule states that, for such a pattern matching a scrutinee already in
WHNF, we introduce the linear components of the pattern as $\D$-bound variables
whose usage environment matches the linear resources required to type the
corresponding constructor argument in the scrutinee, which is annotated in
the judgement ($\ov{\D_i}$). Unrestricted fields of the constructor are
introduced as unrestricted variables. We note that the typing environment $\D$
always contains the resources $\ov{\D_i}$ in uses of the alternative
judgement.

@@

Hence, to type a case expression whose scrutinee is in weak head normal form, we
type the scrutinee with linear resources $\D$ and type the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, annotating the judgement with the proof irrelevant resources,
and using the $\Rrightarrow$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is similar to the one for scrutinees in WHNF, but
requiring the resources in the case binder, typing environment, and
judgement annotation to be made irrelevant.

@@

The key idea to guarantee these is to provide a $\textsc{Split}$ rule that
allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
constructor $K$, where $n$ is the number of linear components of the
constructor, and each resource resulting from the split is \emph{tagged} with
$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
%
By assigning to each pattern variable a \emph{fragment} of the scrutinee
resources (with a tag), we require the scrutinee resources to be
$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
remaining tagged resources cannot be used directly, yet need still be consumed
exactly once. Thus, the choice to consume the tagged resources via usage
environments of the other pattern variables is forced, ensuring that no
variable for a constructor's linear component can go unused (the tagged environments are disjoint).

@@

The list of transformations we have
validated is given in Figure~\ref{fig:opttrans}.
%
The behaviour of most transformations can be inferred from its name:
Inlining substitutes a let-bound expression in the let body; $\beta$-reduction
and case of known constructor effectively perform evaluation of the given expression;
case-of-case commutes cases in scrutinee position inside the outer case alternatives;
let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
the four commuting conversions allow let-bound expressions to be commuted with other
Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
finally, binder swap substitutes occurrences of a scrutinee variable for the
case binder. While many of the transformations above are not obviously optimising,
they are chained in the compiler pipeline to expose optimisation opportunities.

@@

The implementation of Linear Core as a typechecker does not follow
directly from the description of the type system because Linear Core
is not \emph{syntax-directed}. Specifically, the most challenging
features are splitting linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
%
Pattern variables bound in a case alternative, for a scrutinee not in WHNF, are
introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use these pattern variables, the scrutinee resources must be first
$Split$ into fragments. Instead of guessing which resources need to be split,
the implementation consumes tagged fragments of a resource \emph{as needed},
i.e., only when consuming a resource whose usage environment has a resource
tagged $K_j$ do we $Split$ that resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. Further research regarding
this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

@@

The results of running the Linear Core GHC plugin on large 
libraries focused around linear types are given by
Figure~\ref{fig:core-plugin-res}.
%
%
We compiled the Haskell standard library for programming with linear types,
\texttt{linear-base} (4000 lines, comprising over 100 modules), the
\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines) library, and
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
session-types library, using our plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is typechecked once per optimisation pass. i.e., we typecheck
all intermediate programs produced by GHC. The total amount of programs
rejected by the typechecker are counted dually, also accross every optimisation
pass, without halting compilation when a program is rejected. We note that our
implementation is not performance conscious, and types every intermediate
program from scratch, instead of preserving information throughout the
pipeline.

@@

    \item Programs rejected due to the lack of ``linearity coercions'' in
        $\lambda^\pi_\Delta$. So-called \emph{multiplicity coercions} extending
        Linear Core could make these programs be understood as linear.

>>>>

We start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f6 x y use = case (x,y) of z { (a,b) -> use a b }
\end{lstlisting}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is 
\emph{semantically} linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f7 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
We justify that \incode{f7} is linear by appealing to the fact that 
the tuple being scrutinized is already in WHNF and so evaluating it will
not consume either $x$ nor $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we are not using $a$ or $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$
would be unavailable (and so \incode{f8} below cannot be accepted):
\begin{noway}
\begin{lstlisting}
f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f8 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable we run the computations
defined in terms of the linear variables used for that constructor
argument. Otherwise, if we do not use the pattern variables, then we
do not run the computation and thus no resources are consumed.

A third option is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $(a,b)$ and with $(x,y)$, but at least one of
the three must occur to ensure the linear resources are consumed. In this
example, we can think that using $a$ entails using the resource $x$, $b$ the
resource $y$, and the case binder $z$ entails using both $a$ and $b$.
%
Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require performing computation and thus consume linear
resources:
\begin{notyet}
\begin{lstlisting}
f10 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f10 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we must still use either the case binder $z$ or the linear
pattern variables $a,b$. %  e.g. it would be quite disastrous if any of the following typechecked:
% \begin{noway}
% \begin{lstlisting}
% doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
% doubleFree x free = case free x of z { Result v -> free x }
% \end{lstlisting}
% \end{noway}
% % Remover
% \begin{noway}
% \begin{lstlisting}
% leakPointer :: Ptr -> ()
% leakPointer x = case id x of z { _ -> () }
% \end{lstlisting}
% \end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly once to guarantee
that the resources used in the scrutinee are fully consumed. %, or
                                %risk them being only ``almost''
                                %consumed.
For instance, take
\incode{use} in \incode{f10} to be the pairing function \incode{(,)}: it is not sufficient for \incode{use x y} to be
evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, potentially
discarding linear resources. In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

@@

Finally, we discuss case expressions which scrutinize a
single variable $x$, e.g.:
$
\lambda x.~\ccase{x}{\_ \to x}
$. 
It might seem as though this function is linear:
%begin{itemize}
% \item
if the argument instantiating $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact;
% \item
if the argument instantiating$x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
$x$ returns the result of evaluating $x$ that still had to be consumed exactly
once.
%\end{itemize}
However, in practice, this program's linearity can only be determined by
further analysis on the non-strict evaluation strategy. If linear
function applications, in general, are $\beta$-reduced \emph{call-by-name} (a desirable
practice, as linear functions are known to use their argument exactly once)
% and thus call-by-name reduction will definitely not duplicate computation
while the above function is considered linear, then such an application might
duplicate linear resources during evaluation. For instance:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
\Longrightarrow_{CBN}
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
On the other hand, if linear function applications are reduced
\emph{call-by-need}, the above example would instead be semantically linear
because the result of freeing $x$ would be bound to some $y$ first.
%
The system we present in the next section, which evaluates linear
applications call-by-name, does not accept the above function scrutinizing a
variable as linear. 
This interaction between evaluation and linearity arises in the type
preservation proof of our system, and is again explored with the reverse
binder swap transformation discussed in
Section~\ref{sec:optimisations-preserve-types-meta}.

@@

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D_1}\s_1\dots x_n{:}_{\D_n}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. We often omit the case binder $z$ when it is unused. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for type-checking semantic linearity in the presence of laziness we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default} or \emph{wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting whether the pattern-bound variables must
consumed linearly (ultimately, in order to consume the scrutinee linearly).
Additionally, the set of patterns in a case expression is assumed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.

@@

\vspace{-0.5cm}
\[
\TypeVarDelta
\]
The rule reads that an occurrence of a $\Delta$-bound variable is well-typed if
the linear environment is made up exactly of the resources in the usage environment of
that variable.
%
$\D$-variables are unrestricted 
since they can be discarded and duplicated, despite multiple
occurrences of the same $\Delta$-variable not being
well-typed as, ultimately, it would imply non-linear usage of linear
resources.
% multiple occurrences of the same
% $\D$-variable imply non-linear usage of resources that must be used linearly.

@@

Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
a case expression \emph{evaluates its scrutinee} to Weak Head Normal
Form (WHNF)~\cite{10.5555/1096899}, and \emph{then} selects the case
alternative corresponding to the pattern matching the WHNF of the
scrutinee\footnote{In our calculus, the alternatives are assumed to be
  exhaustive, i.e. there always exists at least one pattern which
  matches the scrutinee in its WHNF, so we're guaranteed to have an
  expression to progress evaluation.}. An expression in WHNF can either be %\begin{itemize}
% \item
a $\lambda$-abstraction $\lambda x.~e$ 
or a datatype constructor application $K~\ov{e}$.
%\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments need not be evaluated for the lambda or constructor
application to be in WHNF. % (if otherwise all sub-expressions
% were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
Accordingly, these sub-expressions might still depend on linear resources to be
well-typed (these resources will be consumed when the expressions are evaluated).

As will be made clear in this section, we need to devise a specialized typing
for scrutinees that distinguishes between terms
in WHNF and terms that are not in WHNF.
Following the discussion on expressions in WHNF, we present a
typing judgement $\G;\D \Vdash e : \s \gtrdot \ov{\D_i}$, and a rule for each
of the forms given above:
\[
    \TypeWHNFCons
\qquad
    \TypeWHNFLam
\]
This judgement differs from the main typing judgement in that (1) it only
applies to expressions in weak head normal form, and (2) it ``outputs'' (to the right of $\gtrdot$) a
disjoint set of linear environments ($\ov{\D_i}$), where each environment corresponds to the
linear resources used by a sub-expression of the WHNF expression.
%
To type a constructor application $K~\ov{e_\omega e_i}$, where $e_\omega$
are the unrestricted arguments and $e_i$ the linear arguments of the
constructor, we split the resources $\D$ into a disjoint set of resources
$\ov{\D_i}$ required to type each linear argument individually and return exactly
that split of the resources; the unrestricted $e_\omega$ expressions must be
typed on an empty linear environment. A lambda expression is typed with
the main typing judgement and trivially ``outputs'' the whole $\D$ environment,
as there is always only a single sub-expression in lambda abstractions (the body).

@@

The dichotomy between evaluation (i.e.,~resource usage) of a case expression
whose scrutinee is in weak head normal form, or otherwise, leads to one of our
key insights: we must \emph{branch} on weak head normal formed-ness to
accurately type case expressions.
%
When the scrutinee is already in WHNF, the resources are
unused upon evaluation and thus available in the alternatives.
%
When it is not, resources will be consumed and cannot be used in the
alternative.
%
To illustrate, consider the following case expressions:
\[
\begin{array}{ccc}
(1)~\lambda x.~\ccase{K~x}{\_ \to x} &  & (2)~\lambda x.~\ccase{free~x}{\_ \to x}
\end{array}
\]
The first function uses $x$ linearly, while the second does not.
%
Alternatives may also use the case binder or pattern variables, referring to, respectively,
the whole scrutinee (and all resources used to type the scrutinee) or
constructor arguments (and the resources to type each argument).
% Using the case binder entails using all the resources required by the scrutinee, and using a pattern
% variable implies using the resources of the corresponding constructor argument.

@@

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, encompassing three
``modes'', distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF ($\Mapsto$),
for case expressions in which the scrutinee is not in WHNF ($\Rrightarrow$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t
The rule for typing an alternative whose pattern is a
constructor with $n > 0$ linear components (in mode $\Mapsto$) is:
\[
\TypeAltNWHNF
\]
The rule states that, for such a pattern matching a scrutinee already in
WHNF, we introduce the linear components of the pattern as $\D$-bound variables
whose usage environment matches the linear resources required to type the
corresponding constructor argument in the scrutinee, which is annotated in
the judgement ($\ov{\D_i}$). Unrestricted fields of the constructor are
introduced as unrestricted variables. We note that the typing environment $\D$
always contains the resources $\ov{\D_i}$ in uses of the alternative
judgement.

@@

Hence, to type a case expression whose scrutinee is in WHNF, we
type the scrutinee with linear resources $\D$ and type the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, annotating the judgement with the proof irrelevant resources,
and using the $\Rrightarrow$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is similar to the one for scrutinees in WHNF, but
requiring the resources in the case binder, typing environment, and
judgement annotation to be made irrelevant.

@@

The key idea to guarantee these is to provide a $\textit{Split}$ rule that
allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
constructor $K$, where $n$ is the number of linear components of the
constructor, and each resource resulting from the split is \emph{tagged} with
$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
%
By assigning to each pattern variable a \emph{fragment} of the scrutinee
resources (with a tag), we require the scrutinee resources to be
$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
remaining tagged resources cannot be used directly, yet need still be consumed
exactly once. Thus, the choice to consume the tagged resources via usage
environments of the other pattern variables is forced, ensuring that no
variable for a constructor's linear component can go unused (the tagged environments are disjoint).

@@

The list of transformations we have
validated is given in Figure~\ref{fig:opttrans}.
%
The behaviour of most transformations can be inferred from its name:
Inlining substitutes a let-bound expression in the let body; $\beta$-reduction
and case of known constructor effectively perform evaluation of the given expression;
case-of-case commutes cases in scrutinee position inside the outer case alternatives;
%let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
the five commuting conversions allow let-bound expressions to be commuted with other
Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
finally, binder swap substitutes occurrences of a scrutinee variable for the
case binder. While many of the transformations above are not obviously optimising,
they are chained in the compiler pipeline to expose optimisation opportunities.

@@

The implementation of Linear Core as a typechecker does not follow
directly from the description of the type system because Linear Core
is not \emph{syntax-directed}. Specifically, the most challenging
features are splitting linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
%
Pattern variables bound in a case alternative, for a scrutinee not in WHNF, are
introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use these pattern variables, the scrutinee resources must be first
$Split$ into fragments. Instead of guessing which resources need to be split,
the implementation consumes tagged fragments of a resource \emph{as needed},
i.e., only when consuming a resource whose usage environment has a resource
tagged $K_j$ do we $Split$ that resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. %Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

@@

The results of running the Linear Core GHC plugin on large 
libraries focused around linear types are given by
Figure~\ref{fig:core-plugin-res}.
%
%
We compiled the Haskell standard library for programming with linear types,
\texttt{linear-base} (4000 lines, comprising over 100 modules), the
\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines) library, and
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
session-types library, using our plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is typechecked once per optimisation pass (i.e., we typecheck
all intermediate programs produced by GHC). The total amount of programs
rejected by the typechecker are counted dually, also across every optimisation
pass, without halting compilation when a program is rejected. We note that our
implementation is not performance conscious, and types every intermediate
program from scratch, instead of preserving information throughout the
pipeline.

@@

    \item Programs rejected due to the lack of ``linearity coercions'' in
        $\lambda^\pi_\Delta$. So-called \emph{multiplicity coercions} extending
        Linear Core could make these programs be understood as linear
        (discussed in Section~\ref{sec:future-work}).
